Pareto-Optimal Multistate Protein Design using Tree Graphical Models

### 2012-12-06 docs/tex/draft.tex
Introduction and Conclusion

\begin{abstract}
The problem of protein design is to find a protein sequence which folds into a given protein structure. Multistate protein design generalizes this to find a protein sequence which folds into multiple protein states. This involves finding a sequence that is not only stable in one state but in the other state as well. This problem can be posed as a multi-objective optimization problem. By making suitable modelling approximations using graphical models, I propose to solve this multi-state design problem that allows the discovery of a set of sequences that are pareto optimal. Pareto-optimality in this context refers to the property that it is not possible to improve stability in one state without hurting the stability of the other state. Theory  , experiments and conclusions will be presented. 
\end{abstract}

\section{Introduction}
To be added shortly \cite{Ambroggio2006}.

@@

\section{Conclusion}
Write your conclusion here.

>>>>

\begin{abstract}
The problem of protein design is to find a protein sequence which folds into a given protein structure. Multistate protein design generalizes this to find a protein sequence which folds into multiple protein states. This involves finding a sequence that is not only stable in one state but in the other state as well. This problem can be posed as a multi-objective optimization problem. By making suitable modelling approximations using graphical models we attempt to solve this optimization problem exactly and efficiently. We also show how to efficiently find all pareto-optimal sequences by enumerating the pareto-optimal frontier in sequence space. Thus, we make a novel contribution to the repertoire of algorithms for protein design. 
\end{abstract}

\section{Introduction}
A major goal in medicine is to be able to design biological macromolecules that perform a certain desired biological function. These biological macromolecules called proteins are ubiquitously found in all living organisms. They perform a number of functions ranging from motion, structure, signalling, catalysis, etc. They are the molecular machines of the biological world. One of the primary reasons that proteins are able to perform so many diverse functions is that individual proteins can adopt multiple roles. They do so by being able to show great structural diversity. While it is certainly true that sequence determines structure, it is not true that it is a single static structure. Proteins can adopt multiple structures and conformations. Some proteins are particularly better than others in their ability to adopt distinct conformations. We refer to these proteins as multistate proteins. 
\\
\\
The problem of protein design is to find a protein sequence which folds into a given protein structure. Multistate protein design generalizes this to find a protein sequence which folds into multiple protein states. This involves finding a sequence that is not only stable in one state but in the other state as well. Biologist have previously designed proteins that can switch between folded and disordered states, between distinct folded states and between different aggregation states~\cite{Ambroggio2006}. These proteins switch under a variety of trigger mechanisms such as pH shifts, post-translational modification and ligand binding. However most efforts lead by biologists have heavily relied on detailed prior understanding of the system and trial and error. 
\\
\\
Computational attempts at multistate design have been stymied by the enormous complexity of the protein design space. Pierce et.al~\cite{Pierce2002} proved that protein design is NP-hard. Multistate design a generalization of protein design is by extension NP-hard as well. Previous attempts at multistate protein design have involved a lot of different strategies including Markov Chain Monte Carlo (MCMC)~\cite{Leaver-Fay2011,Ambroggio2006}, Genetic Algorithms, Dead End Elimination~\cite{Yanover2007} and even graphical models~\cite{Fromer2010}.The multistate applications have ranged from negative design, protein-protein binding, homodimerization, allosteric design, etc. Each of these methods make significant contributions to the field of multistate protein design. However each of them suffers from some drawback or the other. MCMC and GA methods are approximate and do not have any guarantees and are expensive. DEE is an exact method, but is very expensive to calculate and cannot be scaled up. BP using MRFs is the most elegant yet, but it also suffers from computational cost and scale-up issues. 
\\
\\
When designing algorithms, there is often a trade-off between accuracy and speed. In this paper, I will present a method that is fast and exact. It is very flexible allowing the user to supply various features. It also allows for full enumeration of the pareto-frontier of the sequence space. Unfortunately, these benefits come at the cost of accuracy and expressive power. Specifically, it makes an sequential dependency assumption between the amino acids of a protein. This is obviously not true since proteins are 3D structures and can fold back in order to interact with non-sequential parts of itself. While it is true that this is a limitation, it remains to be seen how much of a limitation it really is. Many chain methods such as HMMs are very successfully used within the biological community. 
\\
\\
Our method uses tree-structured MRFs(TMRFs)for efficient inference and decoding to find the minimum energy sequence for a particular state. We make a clever observation that minimum energy sequences of convex combinations of states can be found efficiently as well when using a TMRF. This unique property lets us fully enumerate the pareto-frontier of the sequence space thus identifying all the sequences that are strictly dominant to other sequences. Thus it lets us very quickly describe the sequence space. In the rest of the paper we will present the theory behind all the methods. We first introduce Hidden Markov Models (HMMs) for motivation. We then move onto chain structured CMRFs and talk about efficient max-decoding in these frameworks. We further generalize CMRFs to tree structured MRFs (TMRFs) and show how to do max decoding under this setting as well. Then we tie all the models together using the quick hull algorithm for enumerating the sequences in the pareto frontier. Finally, we close by talking about some mathematical subtleties of the proposed method.

@@

\section{Conclusion}
We presented a method that is able to perform fast decoding and inference for finding minimum energy sequences for the purpose of protein design. We also show to find the pareto-frontier for the sequence space. The key idea behind this method is the use of Tree MRFs which allow fast Dynamic Programming for max-decoding.
